openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding

### 2020-10-01 paper/policar2020-opentsne.tex
Edits to main text

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{macosko2015}
\caption{\label{fig:macosko}
Three t-SNE embeddings demonstrate theoretical advances and corresponding implementations in
\opentsne. The data in \textbf{(a)} and \textbf{(b)} represent 44,808
single-cell gene-expression profiles of mouse retinal cells from Macosko
\textit{et al.}~\cite{macosko2015highly}. The data in \textbf{(c)}
additionally contains 27,499 expression profiles from mouse
retinal cells from Shekhar \textit{et
al.}~\cite{shekhar2016comprehensive}. \textbf{(a)} We construct a t-SNE
embedding following the parameter choices from the original publication
by Maaten \& Hinton~\cite{maaten2008visualizing}. The visualization
shows no preservation of the global organization of clusters,
resulting from random initialization and an affinity model focused on
preserving local neighborhoods. \textbf{(b)} A modern t-SNE
embedding, utilizing the latest theoretical advances and practical
recommendations constructed using a multi-scale
affinity model, preserving both short-range and long-range interactions
between data points and initialized so that the global layout is
meaningful. Unlike in \textbf{(a)}, the green and blue clusters
representing different sub-types of amacrine and bipolar cells are now
localized to the same regions of the space, indicating a higher level
of similarity than to other cell types. The embedding in \textbf{(c)} shows how existing
t-SNE reference atlases can be used to place new samples into existing
embeddings. The positions of new data points correspond to cell types
from the reference atlas.
}
\end{figure*}

@@

\begin{figure*}[htbp]
	\includegraphics[width=\textwidth]{tasic2018}
	\caption{\label{fig:tasic}We use \opentsne\ to create four different
	visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data,
	each providing a different perspective into the topology of the data.
	The data set contains 21,874 single-cells originating from the mouse
	neocortex. Cluster annotations and colors are taken from the original
	publication. Warm colors correspond to excitatory neurons, cool colors
	correspond to inhibitory neurons, and gray/brown colors correspond to
	non-neuronal cells. The standard t-SNE \textbf{(a)} emphasizes local
	structure while increasing perplexity \textbf{(b)} results in a more
	meaningful layout of the clusters. We can also combine the two
	perplexities by using a multiscale kernel affinity model \textbf{(c)}
	and obtain a trade-off between global and local structure.
	Alternatively, we can inspect more fine-grained structure and reveal
smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}

@@

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{cao2019}
  \caption{\label{fig:cao}
Increasing the exaggeration parameter leads to compact clusters, highlighting
	the data's global organization and emphasizing the continuity between
	cell states. The data set from Cao \textit{et al.}~\cite{cao2019single}
	contains expression profiles from 2,058,652 single cells. The data were
	collected from mice embryos at different developmental stages at daily
	intervals after 9.5 to 13.5 days. \textbf{(c)} reveals that the data is
	comprised of two main components -- the neural tube and mesenchymal
	cells -- as well as several other smaller clusters. Additionally, the
	colors indicate developmental progression with red indicating
	least-developed cells and blue indicating most developed cells. The
	overall developmental trajectory is most apparent with higher
	exaggeration levels, showing red cells slowly transitioning into blue
	cells. Progressively easing the exaggeration factor uncovers finer
	clusters within the larger groups, as shown in \textbf{(b)} with
	exaggeration of two and subsequently in \textbf{(a)}, where we show the
	standard t-SNE with no exaggeration. 32,011 putative doublets are
	excluded from the visualizations.
}
\end{figure*}

>>>>

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{macosko2015}
  \caption{\label{fig:macosko}
  We use \opentsne\ to generate three t-SNE embeddings and demonstrate recent
  theoretical advances. The data in \textbf{(a)} and \textbf{(b)} represent 44,808
  single-cell gene-expression profiles of mouse retinal cells from Macosko
  \textit{et al.}~\cite{macosko2015highly}. The data in \textbf{(c)}
  additionally contains 27,499 expression profiles from mouse
  retinal cells from Shekhar \textit{et al.}~\cite{shekhar2016comprehensive}.
  \textbf{(a)} We construct a t-SNE
  embedding following the parameter choices from the original publication
  by Maaten \& Hinton~\cite{maaten2008visualizing}. The visualization
  shows no preservation of the global organization of clusters,
  resulting from random initialization and an affinity model focused on
  preserving local neighborhoods. \textbf{(b)} A modern t-SNE
  embedding, utilizing the latest theoretical advances and practical
  recommendations constructed using a multi-scale
  affinity model, preserving both short-range and long-range interactions
  between data points and initialized so that the global layout is
  as meaningful as possible. Unlike in \textbf{(a)}, the green and blue clusters
  representing different sub-types of amacrine and bipolar cells are now
  localized to the same regions of the space, indicating a higher level
  of similarity than to other cell types. The embedding in \textbf{(c)} shows how existing
  t-SNE reference atlases can be used to place new samples into existing
  embeddings. The positions of new data points correspond to cell types
  from the reference atlas.
}
\end{figure*}

@@

\begin{figure*}[htbp]
	\includegraphics[width=\textwidth]{tasic2018}
	\caption{\label{fig:tasic}We use \opentsne\ to create four different
	visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data,
	each providing a different perspective into the topology of the data.
	The data set contains 21,874 single-cells originating from the mouse
	neocortex. Cluster annotations and colors are taken from the original
	publication. Warm colors correspond to excitatory neurons, cool colors
	correspond to inhibitory neurons, and gray/brown colors correspond to
	non-neuronal cells. Standard t-SNE \textbf{(a)} emphasizes local
	structure while increasing perplexity \textbf{(b)} results in a more
	meaningful layout of the clusters. We can also combine the two
	perplexities by using a multiscale kernel affinity model \textbf{(c)}
	and obtain a trade-off between global and local structure.
	Alternatively, we can inspect more fine-grained structure and reveal
  smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}

@@

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{cao2019}
  \caption{\label{fig:cao}
  Increasing the exaggeration parameter leads to compact clusters, highlighting
	the data's global organization and emphasizing the continuous nature of
	cell state transitions. The data set from Cao \textit{et al.}~\cite{cao2019single}
	contains expression profiles from 2,058,652 single cells. The data were
	collected from mice embryos at different developmental stages at daily
	intervals after 9.5 to 13.5 days. \textbf{(c)} reveals that the data is
	comprised of two main components -- the neural tube and mesenchymal
	cells -- as well as several other smaller clusters. The
	colors indicate developmental progression with red indicating
	least-developed cells and blue indicating most developed cells. The
	overall developmental trajectory is most apparent with higher
	exaggeration levels, showing red cells slowly transitioning into blue
	cells. Progressively easing the exaggeration factor uncovers finer
	clusters within the larger groups, as shown in \textbf{(b)} with
	exaggeration of two and subsequently in \textbf{(a)}, where we show the
	standard t-SNE with no exaggeration. 32,011 putative doublets are
	excluded from the visualizations.
}
\end{figure*}
